---
title: 'Reproducible Research: Course Project 2'
author: "Rich Florez"
date: "9/25/2016"
output: html_document
---

##Severe Weather Impacts in the United States from 1950 - 2011##

###Synopsis###

This analysis examines severe weather impact taken from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database as it applies to the United States population's health and economy between the years of 1950 - 2011.

In particular, this data analysis will address the following questions:

1. Across the United States, which types of events (as indicated in the ğ™´ğš…ğšƒğšˆğ™¿ğ™´ variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

As the data analysis below demonstrates:

- **Tornados** are most harmful to *population health* as measured by **fatalities**
- **Flooding** has the greatest *economic impact* as measured by **property damage**

###Data Processing###

The data for this assignment can be downloaded from the course web site:

- [NOAA Storm Data (49Mb zipped/562Mb unzipped)](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

Additional documentation for the data can be found here:

- [Storm Data Documentation (NWS)](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- [FAQ (NCDCSE)](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

Other data processing notes:

- *No data pre-processing was performed on the raw data from the NOAA storm database prior to the analysis*

- *A DamageFactor function was used to transform the property damage data values into comparable figures*

- *The ggplot library was used to generate the plots in the results section*

Below is a summary of the dataset within the relevant working directory indicating 902,297 rows/observations and 37 columns/fields of data:

```{r, echo=TRUE}
setwd("~/datasciencecoursera/ReproRsrch/ProjectWk4/")
stormData <- read.csv("repdata-data-StormData.csv")
dim(stormData)
head(stormData)
```

###Results###

```{r, echo=TRUE}
popFatalSubset <- (subset(stormData, FATALITIES > 0))
popFatal <- aggregate(FATALITIES ~ EVTYPE, data = popFatalSubset, FUN = "sum", na.rm = TRUE)
popFatalDesc <- popFatal[order(popFatal$FATALITIES, decreasing=TRUE), ]
popFatalTop10 <- popFatalDesc[1:10, ]
print(popFatalTop10)
library(ggplot2)
ggplot(popFatalTop10, aes(x=EVTYPE, y=FATALITIES)) +
  geom_bar(stat="identity") +
  xlab("Event Type") + 
  ylab("Fatalities") +
  ggtitle("Top 10 U.S. Fatal Severe Weather Impacts 1950-2011")
```

```{r, echo=TRUE}
DamageFactor <- function(amount, multiplier)
{
  factoredAmount <- 0
  if (toupper(multiplier)[1]=="K")
  {
    factoredAmount <- (amount * 1000)
  }
  if (toupper(multiplier)[1]=="M")
  {
    factoredAmount <- (amount * 1000000)
  }
  if (toupper(multiplier)[1]=="B")
  {
    factoredAmount <- (amount * 1000000000)
  }
  return(factoredAmount)
}
econDamageSubset <- (subset(stormData, PROPDMG > 0))
econDamageSubset$DamageFactor <- mapply(DamageFactor, econDamageSubset$PROPDMG, econDamageSubset$PROPDMGEXP)
econDamage <- aggregate(DamageFactor ~ EVTYPE, data = econDamageSubset, FUN = "sum", na.rm = TRUE)
econDamageDesc <- econDamage[order(econDamage$DamageFactor, decreasing=TRUE), ]
econDamageTop10 <- econDamageDesc[1:10, ]
print(econDamageTop10)
library(ggplot2)
ggplot(econDamageTop10, aes(x=EVTYPE, y=DamageFactor)) +
  geom_bar(stat="identity") +
  xlab("Event Type") + 
  ylab("Property Damage in US$") +
  ggtitle("Top 10 U.S. Economic Severe Weather Impacts 1950-2011")
```

To reiterate the conclusions first mentioned in the synopsis, the data analysis above demonstrates:

- **Tornados** are most harmful to *population health* as measured by **fatalities**
- **Flooding** has the greatest *economic impact* as measured by **property damage**